Light inextensible strings under tension in the Schwarzschild geometry

### 2019-08-26 string_physics_teacher.tex
initial work, emphasising the educational nature of the results

Light inextensible strin
gs are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
$c^4/\left(4GM\right)$ for the strength of gravity at the event
horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
c^4\right)$.  For the string to behave as though light on lengthscales
comparable with the Schwarzschild radius would require $T\gtrsim\mu
c^2$.  Exceeding this criterion would violate the dominant energy
condition (and in this context we note that transverse waves would
propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
idealization of a string as light and inextensible is an interesting
and informative limit: Rindler~\cite{rindler}, for example, considers
lowering a heavy object to the event horizon on the end of such a
string to illustrate energetic arguments in GR.

>>>>

General relativity is a topic of long-standing fascination for
undergraduate physics students [CITATION NEEDED].  Einstein's concept
of ``curved'' spacetime is an enduring metaphor that attracts large
numbers of undergraduates into courses such as AUT's frontiers of
astronomy courses.

Students are weary of being told that ``space is warped'' and ``not
even light can escape a black hole''.  Students typically stumble on
the concept of warped spacetime on the grounds that torsion-free
metrics are not embedded IN ANYTHING.

Here I show how numerical methods can be used to interpret the concept
of ``straight line'' in the vicinity of a black hole.  As
\cite{overduin2018} point out, demonstrations of relativistic concepts
using everyday materials (in their case gravitational waves visualised
by oscillations in spandex) can exert a galvanising effect on
students.

Light inextensible strings are a stalwart feature of elementary
physics.  It is perhaps not out of place to point out explicitly that
a light string is not subject to gravity, and its being under tension
means that its configuration is an extremum of proper length.  Of
course, no real string is perfectly weightless or inextensible.  With
tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
$c^4/\left(4GM\right)$ for the strength of gravity at the event
horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
c^4\right)$.  For the string to behave as though light on lengthscales
comparable with the Schwarzschild radius would require $T\gtrsim\mu
c^2$.  Exceeding this criterion would violate the dominant energy
condition (and in this context we note that transverse waves would
propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
idealization of a string as light and inextensible is an interesting
and informative limit: Rindler~\cite{rindler}, for example, considers
lowering a heavy object to the event horizon on the end of such a
string to illustrate energetic arguments in GR.
